---
title: "EDLD 651 Final"
author: "Alex Newson, Deanna Strayer, Sabreen NoorAli, Seulbi Lee, Tony Daza"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_folding: hide
---

```{r setup, include=FALSE}
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(ggridges)
library(kableExtra)
library(psych)
library(stats)
library(ggstatsplot)
```

# Research Question
**Do the teacher's reported student math scores across the two timepoints (i.e., base year and first follow-up) correlate with 1) students' sex, 2) students' race/ethnicity, and 3) mother's highest education level of students?**

# Data Information
## Variables
### Independent Variables: 
##### 1.[BYSEX]
#######  1 = "Male"
#######  2 = "Female"
####### -4 = "Nonrespondent"
####### -8 = "Survey component legitimate skip/NA"
##### 2. [BYRACE]
#######  1 = "Amer. Indian/Alaska Native, non-Hispanic"
#######  2 = "Asian, Hawaii/Pac. Islander,non-Hispanic"
#######  3 = "Black or African American, non-Hispanic"
#######  4 = "Hispanic, no race specified"
#######  5 = "Hispanic, race specified"
#######  6 = "More than one race, non-Hispanic"
#######  7 = "White, non-Hispanic"
#######  -4 = "Nonrespondent"
#######  -8 = "Survey component legitimate skip/NA"
##### 3. [BYMOTHED]
#######  1 = "Did not finish high school"
#######  2 = "Graduated from high school or GED"
#######  3 = "Attended 2-year school, no degree"
#######  4 = "Graduated from 2-year school"
#######  5 = "Attended college, no 4-year degree"
#######  6 = "Graduated from college"
#######  7 = "Completed Master's degree or equivalent"
#######  8 = "Completed PhD, MD, other advanced degree"
#######  -4 = "Nonrespondent"
#######  -8 = "Survey component legitimate skip/NA"
#######  -9 = "Missing"

### Dependent Variables: 
##### 1. [BYTXMSTD] Math test standardized score 
####### Description: Math standardized T Score. The standardized T score provides a norm-referenced measurement of achievement, that is, an estimate of achievement relative to the population (spring 2002 10th-graders) as a whole. It provides information on status compared to peers (as distinguished from the IRT-estimated number-right score which represents status with respect to achievement on a particular criterion set of test items). The standardized T score is a transformation of the IRT theta (ability) estimate, rescaled to a mean of 50 and standard deviation of 10.
##### 2. [F1TXMSTD] F1 math test standardized score 
####### Description: Math standardized T Score. The standardized T score provides a norm-referenced measurement of achievement, that is, an estimate of achievement relative to the population (spring 2004 12th-graders) as a whole. It provides information on status compared with peers (as distinguished from the IRT-estimated number-right score which represents status with respect to achievement on a particular criterion set of test items). Although the T score is reported for all F1 in-school responding students (including transfer students), regardless of grade level, the comparison group for standardizing is the 12th grade population. The standardized T score is a transformation of the IRT theta (ability) estimate, and has a mean of 50 and standard deviation of 10 for the weighted subset of 12th-graders in the sample.

# Data Cleaning 

```{r data cleaning, message=FALSE}
#retrieve data
#els <- read_csv("./data/els_02_12_byf3pststu_v1_0.csv")
#select columns 
#els <- els %>% dplyr::select(STU_ID, BYSEX, BYRACE, BYMOTHED, BYTXMSTD, F1TXMSTD)
#save the revised (cleaned) data to csv
#write.csv(els,"./data/els_cleaned.csv", row.names = FALSE)

els <- read_csv("./data/els_cleaned.csv")

#replace missing data code to NA
els$BYSEX <- na_if(els$BYSEX, -4)
els$BYSEX <- na_if(els$BYSEX, -8)
els$BYRACE <- na_if(els$BYRACE, -4)
els$BYRACE <- na_if(els$BYRACE, -8)
els$BYMOTHED <- na_if(els$BYMOTHED, -4)
els$BYMOTHED <- na_if(els$BYMOTHED, -8)
els$BYMOTHED <- na_if(els$BYMOTHED, -9)
els$BYTXMSTD <- na_if(els$BYTXMSTD, -8)
els$F1TXMSTD <- na_if(els$F1TXMSTD, -8)

#remove if the row doesn't have both BY and F1 math scores
els <- els %>% 
  filter(!is.na(BYTXMSTD) | !is.na(F1TXMSTD))

#rename
els <- els %>%
  mutate(BYSEX = recode(BYSEX, 
                          `1` = "Male", 
                          `2` = "Female"),
         BYRACE = recode(BYRACE, 
                          `1` = "Native American/Alaskan", 
                          `2` = "Asian", 
                          `3` = "Black", 
                          `4` = "Hispanic (no race specified)", 
                          `5` = "Hispanic (specified)", 
                          `6` = "More than one race, non-Hispanic", 
                          `7` = "White, non-Hispanic"),
         BYMOTHED = recode(BYMOTHED,
                          `1` = "Did not finish high school",
                          `2` = "Graduated from high school or GED",
                          `3` = "Attended 2-year school, no degree",
                          `4` = "Graduated from 2-year school",
                          `5` = "Attended college, no 4-year degree",
                          `6` = "Graduated from college",
                          `7` = "Completed Master's degree or equivalent",
                          `8` = "Completed PhD, MD, other advanced degree"))


#rename columns to use pivot_longer
colnames(els)[colnames(els) %in% c("BYTXMSTD", "F1TXMSTD")] <- c("Base", "Follow-up")

els_longer <- els %>% 
  pivot_longer(
    cols = c('Base', 'Follow-up'),
    names_to = "YEAR",
    values_to = "MATH"
  )

els_wider_by <- els %>% 
  pivot_wider(
    id_cols = !'Follow-up',
    names_from = BYRACE,
    values_from = c(Base)
  )

els_wider_f1 <- els %>% 
  pivot_wider(
    id_cols = !Base,
    names_from = BYRACE,
    values_from = c('Follow-up')
  )

```
# Visualization 1

```{r visualization 1, message=FALSE, warning=FALSE}
vis1Data <- els_longer %>% 
  mutate(YEAR = factor(YEAR,
                       levels = c("Follow-up",
                                 "Base"))) %>%
  filter(!is.na(BYSEX)) %>%
  ggplot(aes(x=MATH,y=YEAR,fill=YEAR)) +
    geom_col(position="dodge", show.legend = FALSE) +
    facet_wrap(~ BYSEX,ncol=1) +
    labs(x="Math Scores",
         y="Year",
         title="Student Math Scores",
         subtitle="by year and sex"
         ) +
    scale_fill_manual(values = c("maroon", "gold")) +
  theme_light()
vis1Data
```

# Visualization 2

```{r visualization 2, message=FALSE, warning=FALSE}
vis2Data <- els_longer %>% 
  mutate(YEAR = factor(YEAR,
                       levels = c("Follow-up",
                                  "Base"))) %>%
  filter(!is.na(BYRACE)) %>%
  ggplot(aes(x=MATH,y=YEAR,fill=YEAR)) +
    geom_col(position="dodge", show.legend = FALSE) +
    facet_wrap(~ BYRACE,ncol=1) +
    labs(x="Math Scores",
         y="Year",
         title="Student Math Scores",
         subtitle="by year and race"
         ) +
    scale_fill_manual(values = c("maroon", "gold")) +
  theme_light()
vis2Data

# Alternate graph combining Visualization 1 & 2? Maybe easier that way?

vis2DataAlternate <- els_longer %>% 
  mutate(YEAR = factor(YEAR,
                       levels = c("Follow-up",
                                  "Base"))) %>%
  mutate(BYSEX = factor(BYSEX,
                       levels = c("Male",
                                  "Female"))) %>%
  filter(!is.na(BYRACE)) %>%
  ggplot(aes(x=MATH,y=YEAR,fill=BYSEX)) +
    geom_col(position="dodge") +
    facet_wrap(~ BYRACE,ncol=1) +
    scale_fill_discrete(breaks=c('Male', 'Female')) +
    labs(x="Math Scores",
         y="Year",
         fill = "Sex",
         title="Student Math Scores",
         subtitle="by year and race, separated by sex"
         ) +
    scale_fill_manual(values = c("maroon", "gold")) +
  theme_light()
vis2DataAlternate
```

# Visualization 3

## Distribution plots
These are simple distribution plots by Race, Sex, and Mother's education for year 1 and year 2 for standardized math scores.

```{r visualization 3, message=FALSE, warning=FALSE}
# Fixed the names to look better.
els_viz <- els_longer %>%
  mutate(RACE = recode(BYRACE, "Native American/Alaskan" = "Native American\n /Alaskan",
                         "Asian" = "Asian", 
                         "Black" = "Black", 
                         "Hispanic (no race specified)" = "Hispanic", 
                         "Hispanic (specified)" = "Hispanic\n (Race specified)", 
                         "More than one race, non-Hispanic" = "2+ races\n non-Hispanic",
                         "White, non-Hispanic" = "White\n non-Hispanic"),
         MOTHED = recode(BYMOTHED, 
                           "Did not finish high school" = "Did not finish\n high school",
                           "Graduated from high school or GED" = "Graduated high\n school or GED",
                           "Attended 2-year school, no degree" = "Attended 2-year school\n no degree",
                           "Graduated from 2-year school" = "Graduated 2-year\n school",
                           "Attended college, no 4-year degree" = "Attended college\n no degree",
                           "Graduated from college" = "Graduated college",
                           "Completed Master's degree or equivalent" = "Master's degree",
                           "Completed PhD, MD, other advanced degree" = "PhD, MD,other\nadvanced degree")) %>%
  mutate(RACE = factor(RACE, levels = c("White\n non-Hispanic",
                                        "Black",
                                        "Hispanic",
                                        "Hispanic\n (Race specified)",
                                        "Asian",
                                        "Native American\n /Alaskan",
                                        "2+ races\n non-Hispanic")),
         MOTHED = factor(MOTHED, levels = c("Did not finish\n high school",
                                            "Graduated high\n school or GED",
                                            "Attended 2-year school\n no degree",
                                            "Graduated 2-year\n school",
                                            "Attended college\n no degree",
                                            "Graduated college",
                                            "Master's degree",
                                            "PhD, MD,other\nadvanced degree")))


# A plot of the distribution of math scores by race in year 1 and follow up  
els_viz %>%
  filter(!is.na(MATH) & !is.na(RACE)) %>%
  ggplot(aes(x = MATH)) +
  geom_histogram(col='black',fill='white')+
  theme_minimal() +
  xlab("Math Scores") +
  xlim(10,90)+
  facet_wrap( ~ RACE + YEAR, nrow = 2, ncol=7)+
  theme(strip.background =element_rect(fill="white"))

# A plot of the distribution of math scores by sex in year 1 and follow up.
els_viz %>%
  filter(!is.na(MATH) & !is.na(RACE) & !is.na(BYSEX)) %>%
  ggplot(aes(x = MATH)) +
  geom_histogram(col='black',fill='white')+
  theme_minimal() +
  xlab("Math Scores") +
  xlim(10,90)+
  facet_wrap( ~ BYSEX + YEAR, nrow = 1, ncol=4)+
  theme(strip.background =element_rect(fill="white"))

# A Plot of distribution of math scores by mother's education for year 1 and follow up. 
els_viz %>%
  filter(!is.na(MATH) & !is.na(MOTHED)) %>%
  ggplot(aes(x = MATH)) +
  geom_histogram(col='black',fill='white')+
  theme_minimal() +
  xlab("Math Scores") +
  xlim(10,90)+
  facet_wrap( ~ MOTHED + YEAR, nrow = 2, ncol=8)+
  theme(strip.background =element_rect(fill="white"))

# Honestly I don't love the visualization, so let's look at a box plot of the data.

# How about a density plot?
```

## Boxplots
A boxplot of the standardized math scores by Race, Sex, and Mother's education.  

```{r visualization 4, message=FALSE, warning=FALSE}
# Boxplot of math scores by Race and Sex separated by Year
els_viz %>%
  filter(!is.na(MATH) & !is.na(RACE)) %>%
  ggplot(aes(x= RACE, y=MATH)) +
  geom_boxplot(aes(fill = RACE), show.legend = FALSE)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_viridis_d(option = 'plasma')+
  theme_minimal()+
  facet_wrap(~BYSEX)+
  labs(x = "", 
       y = "Math Scores",
       title = "Math Score by Race and Year",
       subtitle = "Separated by sex")+
  coord_flip()

# Boxplot of math scores by Mother's education separated by Year
els_viz %>%
  filter(!is.na(MATH) & !is.na(MOTHED)) %>%
  ggplot(aes(x= MOTHED, y=MATH)) +
  geom_boxplot(aes(fill = MOTHED), show.legend = FALSE)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  coord_flip()+
  scale_fill_viridis_d(option = 'mako')+
  theme_minimal()+
  facet_wrap(~YEAR)+
  labs(x = "", 
       y = "Math Scores",
       title = "Math Score by Mother Education and Year")

# Boxplot of math scores by sex separated by year
els_viz %>%
  filter(!is.na(MATH) & !is.na(BYSEX)) %>%
  ggplot(aes(x= BYSEX, y=MATH)) +
  geom_boxplot(aes(fill = BYSEX), show.legend = FALSE)+
  scale_fill_viridis_d()+
  theme_minimal()+
  facet_wrap(~YEAR)+
  labs(x = "", 
       y = "Math Scores",
       title = "Math Score by Sex and Year")
```

## Density Plots

Finally let's examine the data using density plots.
```{r visualization 5, message=FALSE, warning=FALSE}
# Density plot of math scores by Race and Sex.
els_viz %>%
  filter(!is.na(MATH) & !is.na(RACE)) %>%
  ggplot(aes(x = MATH, y = RACE))+
  geom_density_ridges(aes(fill = RACE), alpha=0.5)+
  scale_fill_viridis_d(option = 'plasma')+
  theme_minimal()+
  theme(legend.position = "none")+
  facet_wrap(~BYSEX)

# Denisty plot of math scores by Mother education
els_viz %>%
  filter(!is.na(MATH) & !is.na(MOTHED)) %>%
  ggplot(aes(x = MATH, y = MOTHED))+
  geom_density_ridges(aes(fill = MOTHED), alpha=0.5)+
  scale_fill_viridis_d()+
  theme_minimal()+
  theme(legend.position = "none")+
  labs(x = "Math Score",
       y = "Mother's Education Level")

# Density plot of math scores by mother education separated by year
els_viz %>%
  filter(!is.na(MATH) & !is.na(MOTHED)) %>%
  ggplot(aes(x = MATH, y = MOTHED))+
  geom_density_ridges(aes(fill = MOTHED), alpha=0.5)+
  scale_fill_viridis_d()+
  theme_minimal()+
  theme(legend.position = "none")+
  labs(x = "Math Score",
       y = "Mother's Education Level")+
  facet_wrap(~BYSEX, nrow = 1)

```

# Descriptive Statistics

```{r descriptive, message=FALSE}
By_Race <- els_longer %>%
  group_by(BYRACE) %>%
  summarize(race_n = n(), 
            mean_math = mean(MATH, na.rm = TRUE), 
           sd_math = sd(MATH, na.rm = TRUE))

describeBy(els_longer$MATH, els_longer$BYRACE)

By_Race %>%
  kbl(caption = "Descriptives by Race", digits = 2) %>%
  kable_classic()

By_MotherED <- els_longer %>%
  group_by(BYMOTHED) %>%
  summarize(mothed_n = n(), 
            mean_math = mean(MATH, na.rm = TRUE), 
            sd_math = sd(MATH, na.rm = TRUE))

describeBy(els_longer$MATH, els_longer$BYMOTHED)

By_MotherED %>%
  kbl(caption = "Descriptives by Mother's Education Level", digits = 2) %>%
  kable_classic()
            
By_Year <- els_longer %>%
  group_by(YEAR) %>%
  summarize(year_n = n(), 
            mean_math = mean(MATH, na.rm = TRUE), 
            sd_math = sd(MATH, na.rm = TRUE))

describeBy(els_longer$MATH, els_longer$YEAR)


By_Year %>%
  kbl(caption = "Descriptives by Year", digits = 2) %>%
  kable_classic()

By_Sex <- els_longer %>%
  group_by(BYSEX) %>%
  summarize(sex_n = n(), 
            mean_math = mean(MATH, na.rm = TRUE), 
            sd_math = sd(MATH, na.rm = TRUE))

 describeBy(els_longer$MATH, els_longer$BYSEX)

By_Sex %>%
  kbl(caption = "Descriptives by Sex", digits = 2) %>%
  kable_classic()
```

# More Statistics
## ANOVA Results
### Math scores by Mother Education and Race

```{r ANOVA, warning=FALSE, message=FALSE}
library(car)
math_mod <- lm(MATH ~ 1 + MOTHED*RACE, data = els_viz)
Anova(math_mod, type = 3)
```

## Student's T test Analysis
### Difference in scores between year 1-2

```{r t_test}
#str(els_viz)
# Paired student's t test to examine if means from year 1 and year 2 are significantly different

# Created a small data set of just years and scores plus student id
els_byyear <- els_viz %>%
  group_by(YEAR)%>%
  filter(!is.na(MATH) | !is.na(YEAR)) %>%
  select(STU_ID, YEAR, MATH)%>%
  pivot_wider(names_from = YEAR,
              values_from = MATH) %>%
  rename("Follow" = "Follow-up")

# Paired student's t test to examine if the mean from the follow up is significantly greater than the base year mean
t.test(els_byyear$Follow, els_byyear$Base, paired = TRUE, alternative = "greater")

# Paired student's t test to examine if means from base year and follow up are significantly different.
t.test(els_byyear$Base, els_byyear$Follow, paired = TRUE)

# Visualization of the means by year. The t test suggests that there is a significant difference in scores by year, despite the means being similar, unfortunately, the mean of the follow up is significantly less than the base year. I was thinking we could focus on visualizing the data by race and mother's education and sex rather than comparing year 1 to year 2 data.
ggwithinstats(data = els_viz, x = YEAR, y = MATH, 
              type = "parametric", 
              centrally.plotting = TRUE, 
              pairwise.display = "s", 
              point.path = FALSE,
              point.args = aes(size = 0, alpha = 0.2),
              results.subtitle = FALSE,
              alternative = "greater")

```

## Regression Analysis
```{r regression, warning=FALSE}
# First let us look if Race is a predictor. I will set white as the reference group since it is the largest group. 
#contrasts(els_viz$RACE)
# Looks like I set white as the reference earlier.  

mod_race <- lm(MATH ~ 1 + RACE, els_viz)
summary(mod_race)
# This just tells us that all groups are significantly different from each other
# I think we could report pairwise comparisons

race_pairs <- pairwise.t.test(els_viz$MATH, els_viz$RACE, p.adjust.method = "bonf")
race_pval <- race_pairs$p.value %>%
  round(digits = 3)

options(knitr.kable.NA = "")
race_pval %>%
  kbl(caption = "p -values of Math Score by Race",
      digits = 3) %>%
  kable_classic()
  
# Based on this, most groups are significantly different **except** for black/Native, Hispanic/Hispanic (Race specified), Native/Hispanic, Native/Hispanic (Race specified)

aovRace <-  aov(MATH ~ RACE,data=els_viz)
tukey_race <- TukeyHSD(aovRace,conf.level = 0.95)
tukey_race

# Let's look if Mother's education is a predictor. Did not finish high school is the reference group
#contrasts(els_viz$MOTHED)

mod_mothed <- lm(MATH ~ 1 + MOTHED, els_viz)
summary(mod_mothed)
# This just tells us that groups are significantly different from each other
# I think we could report pairwise comparisons

mothed_pairs <- pairwise.t.test(els_viz$MATH, els_viz$MOTHED, p.adjust.method = "bonf")
mothed_pval <- mothed_pairs$p.value %>%
  round(digits = 3)

options(knitr.kable.NA = "")
mothed_pval %>%
  kbl(caption = "p -values of Math Score by Mother's Education level",
      digits = 3) %>%
  kable_classic()


```


## Mother Education / Race Regression  
I am not sure how to do a regression analysis of all three variables, but I can look at the regression analysis of math scores based on race/ethnicity and mother's education.   

```{r regression mother edu and race}
mod_moth_race <- lm(MATH ~ 1 + MOTHERED*RACE, data = els_viz)
summary(mod_moth_race)

```

### Mother Edu / Race Regression Visualization  
Now let's try to visualize this, it might get messy.  

```{r}
# Visualization of regression models for mother's education and race
#install.packages("interactions")
library(interactions)
cat_plot(mod_moth_race, pred = MOTHED, modx = RACE, geom = "line", interval = FALSE,vary.lty = TRUE)

```

