---
title: "EDLD 651 Final"
author: "Alex Newson, Deanna Strayer, Sabreen NoorAli, Seulbi Lee, Tony Daza"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(readr)

#retrieve data
#els <- read_csv("./els_02_12_byf3pststu_v1_0.csv")
#select columns 
#els <- els %>% dplyr::select(STU_ID, BYSEX, BYRACE, BYMOTHED, BYWRTNGA)
#save the revised (cleaned) data to csv
#write.csv(els,"./els_cleaned.csv", row.names = FALSE)

els <- read_csv("./els_cleaned.csv")

```

### Research Question
##### Do the teacher's reported student writing ability correlate with 1) students' sex, 2) students' race/ethnicity, and 3) mother's highest education level of students?

### Data [need to be rephrased]
##### ELS:2002 represents a major longitudinal effort designed to provide trend data about critical transitions experienced by students as they proceed through high school and into postsecondary education or their careers. The 2002 sophomore cohort is being followed, initially at 2-year intervals, to collect policy-relevant data about educational processes and outcomes. These data pertain especially to student learning, predictors of dropping out, and high school correlates of students’ access to and persistence and attainment in postsecondary education, and their entry into the workforce. 

### Independent Variables: 
##### 1.[BYSEX]
#######  1 = "Male"
#######  2 = "Female"
####### -4 = "Nonrespondent"
####### -8 = "Survey component legitimate skip/NA"
##### 2. [BYRACE]
#######  1 = "Amer. Indian/Alaska Native, non-Hispanic"
#######  2 = "Asian, Hawaii/Pac. Islander,non-Hispanic"
#######  3 = "Black or African American, non-Hispanic"
#######  4 = "Hispanic, no race specified"
#######  5 = "Hispanic, race specified"
#######  6 = "More than one race, non-Hispanic"
####### 7 = "White, non-Hispanic"
#######  -4 = "Nonrespondent"
#######  -8 = "Survey component legitimate skip/NA"
##### 3. [BYMOTHED]
#######  1 = "Did not finish high school"
#######  2 = "Graduated from high school or GED"
#######  3 = "Attended 2-year school, no degree"
#######  4 = "Graduated from 2-year school"
#######  5 = "Attended college, no 4-year degree"
#######  6 = "Graduated from college"
#######  7 = "Completed Master's degree or equivalent"
#######  8 = "Completed PhD, MD, other advanced degree"
#######  -4 = "Nonrespondent"
#######  -8 = "Survey component legitimate skip/NA"
#######  -9 = "Missing"
### Dependent Variables: 
##### [BYWRTNGA] Student writing ability (teacher-reported)
####### Description: This variable is a scale of student writing ability as reported by the student's base-year English teacher. Higher values represent higher teacher-rated student writing ability. Variable was created through principal factor analysis (weighted by BYSTUWT) and standardized to a mean of 0 and standard deviation of 1. Teacher responses about student writing ability were reverse-coded (so higher values on variables would indicate greater student writing ability), and only students for whom the English teacher provided a full set of responses were assigned a scale value. The coefficient of reliability (alpha) for the scale is 0.95.
####### Note: This composite is constructed from English Teacher Questionnaire variables; thus, values of -4 (nonrespondent) and -8 (survey component legitimate skip) refer to missing data from the English Teacher Questionnaire.


### Data Cleaning 

```{r data cleaning}

# final requirement: pivot_longer(), mutate (), select(), filter (), pivot_wider()


#



```


```{r visualization 1}

```


```{r visualization 2}

```


```{r descriptive}

```
