---
title: "EDLD 651 Final"
author: "Alex Newson, Deanna Strayer, Sabreen NoorAli, Seulbi Lee, Tony Daza"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    code_folding: hide
---

```{r setup, include=FALSE}
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(ggridges)
library(kableExtra)
library(psych)
library(stats)
library(ggstatsplot)
#when installing say no to the prompt 
library(multcompView)
# visualizing the Tukeys pairwise comparison
#library(sjPlot)
#library(sjmisc)
#library(sjlabelled)
#These three are for the regression table to make it easier to read
library(interactions)
#This is for the graph of the interaction plot between race and mother's education
```
# Correlations Between Student Variables and Math Results Over Time: An Exploratory Study

## Introduction
#### The data set that we chose comes from the IES National Center for Education Statistics. The specific study is titled *Education Longitudinal Study of 2002* or ELS:2002.
#### ELS:2002 represents a major longitudinal effort designed to provide trend data about critical transitions experienced by students as they proceed through high school and into postsecondary education. The 2002 sophomore cohort was followed, initially at 2-year intervals, to collect policy-relevant data about educational processes and outcomes. These data focus on areas of student learning, student tragectories, student persistence and access to college, as well as entry into the workforce.
#### The baseline year for the study was Spring term 2002. A national sample of high school sophomores were surveyed along with their parents, teachers, adminstrators, and librarians. The first follow-up took place two years later in 2004. Students who remained in the same school for both 2002 and 2004 were resurveyed and tested in multiple areas including mathmatics. Students that did not attend the same school for any reason (i.e. transferred, dropped out, early graduation) were adminstered a questionnaire. Demographic information was collected for both groups along with academic scores.

#### The aim of our project was to  discern the following: **Do the teacher's reported student math scores across the two timepoints (i.e., base year and first follow-up) correlate with 1) students' sex, 2) students' race/ethnicity, and 3) mother's highest education level of students?**

## Methods
#### The National Center for Education Statistics (NCES, n.d.-a) sponsored a longitudinal study to collect data from interviews with 10th grade students in 2002 and three follow-ups in 2004, 2006, and 2012 and their transcripts from high school and postsecondary education in 2005 and 2013. Data used for this project was extracted from the NCES Online Codebook and only included the base year and the first follow-up data (NCES, n.d.-b). Out of approximately 17,600 sophomores who were eligible for the base-year interviews, about 15,400 (87%) students completed a base-year questionnaire in 2002 (Bozick & Lauff, 2007). Around 7,100 of their teachers participated in completing questionnaires, out of the 750 participating schools in the United States (Bozick & Lauff, 2007).  For the first follow-up questionnaire, about 15,000 seniors in high school out of 16,500 eligible students (89%) participated in 2004 (Bozick & Lauff, 2007). 
### Procedures
#### Five variables (i.e., BYSEX, BYRACE, BYMOTHED, BYTXMSTD, F1TXMSTD) out of 4,012 variables included in this longitudinal study were filtered to answer the research question: Does the students’ standardized math scores across the two timepoints (i.e., base year and first follow-up) correlate with 1) students' sex, 2) students' race/ethnicity, and 3) mother's highest education level of students? As independent variables, students’ sex, race, and their mothers’ highest education were filtered. The first dependent variable, BYTXMSTD, is math standardized T scores that were measured in spring 2002 from 10th students. The second dependent variable, F1TXMSTD, is the math standardized T score collected in spring 2004 from 12th students. The standardized T scores in both variables are a transformation of the IRT theta estimate with a mean of 50 and a standard deviation of 10 (NCES, n.d.-b). The raw data were cleaned using the following R functions first: mutate(), select(), filter(), pivot_longer(), and pivot_wider(). Using the cleaned data and the R function of ggplot, a distribution plot, a boxplot, and a density plot were created to visualize the data, gain preliminary information about the distribution, and demonstrate results. The data was subsequently analyzed using two-way analysis of variance (ANOVA) along with post-hoc comparison using tukey’s HSD and a multiple regression analysis of race and mother’s education to answer the research questions. 

## Results
#### First, a student’s t test was conducted to examine if the math scores for the follow up year were significantly different than the math scores from the base year. The paired-student’s t-test resulted in t (13393) = 20.54, p < 0.001 with a mean difference of 0.84 between the base year and follow up year scores. The results suggest that the follow up year had significantly lower math scores than the base year. Next, a two-way analysis of variance (ANOVA) was conducted to examine the extent to which math scores depended on mother’s education and race. The main effect for race was significant f (1, 6) = 57.5, p < 0.0001. Additionally, the main effect for mother’s education was significant f (1, 7) = 270.2, p < 0.0001. The interaction between race and mother’s education for math score was significant f (2, 42) = 4.75, p < 0.0001. Due to the significant interaction effect, post-hoc pairwise comparisons were examined for single one-way ANOVA analysis of race and math as well as mother’s education and math using tukey’s HSD method for pairwise comparison. Nearly all of the comparisons for racial group and mother’s education were significant. The only pairings that were not significant were PhD, MD, or other advanced degree compared to master’s degree (p = .151), PhD, MD, or other advanced degree compared to graduated from 4 year college (p = 0.409), or attended college but did not graduate compared to graduated 2 year school (p = 1). For the race groups, only Native American/Alaskan compared to Hispanic (Race specified and unspecified) (p = 1), as Native American compared to black (p = 0.08), and Hispanic race specified compared to Hispanic race unspecified (p = 1). 
#### A regression analysis was conducted to examine the math score results using race and mother’s education as predictors with the reference group being students whose mothers’ did not graduate from high school for the pairs and the reference for race being white students. Across the board, when race was held constant, student’s whose mother’s had higher levels of education were predicted to score significantly higher in math. Additionally, when mother’s education was held constant, white students were predicted to score significantly higher than black and Hispanic students and significantly lower than Asian students. Examining the pairs, black student scores were predicted to be significantly lower compared to white students at all levels of mother’s education except for the master’s degree level, p = 0.052. The predicted difference between Hispanic student math scores for both race specified and race unspecified did not have as consistent of a pattern, although scores were consistently predicted to be lower than white student scores, the difference was only significant for mother’s education of graduated high school, attend 2-year school, and graduated college for Hispanic race unspecified. The R2 is 0.214 suggesting that 21.4 percent of the change in math scores can be explained by race and mother’s education.  

```{r data_information}
# Data Information
## Variables
### Independent Variables: 
##### 1.[BYSEX]
#######  1 = "Male"
#######  2 = "Female"
####### -4 = "Nonrespondent"
####### -8 = "Survey component legitimate skip/NA"
##### 2. [BYRACE]
#######  1 = "Amer. Indian/Alaska Native, non-Hispanic"
#######  2 = "Asian, Hawaii/Pac. Islander,non-Hispanic"
#######  3 = "Black or African American, non-Hispanic"
#######  4 = "Hispanic, no race specified"
#######  5 = "Hispanic, race specified"
#######  6 = "More than one race, non-Hispanic"
#######  7 = "White, non-Hispanic"
#######  -4 = "Nonrespondent"
#######  -8 = "Survey component legitimate skip/NA"
##### 3. [BYMOTHED]
#######  1 = "Did not finish high school"
#######  2 = "Graduated from high school or GED"
#######  3 = "Attended 2-year school, no degree"
#######  4 = "Graduated from 2-year school"
#######  5 = "Attended college, no 4-year degree"
#######  6 = "Graduated from college"
#######  7 = "Completed Master's degree or equivalent"
#######  8 = "Completed PhD, MD, other advanced degree"
#######  -4 = "Nonrespondent"
#######  -8 = "Survey component legitimate skip/NA"
#######  -9 = "Missing"

### Dependent Variables: 
##### 1. [BYTXMSTD] Math test standardized score 
####### Description: Math standardized T Score. The standardized T score provides a norm-referenced measurement of achievement, that is, an estimate of achievement relative to the population (spring 2002 10th-graders) as a whole. It provides information on status compared to peers (as distinguished from the IRT-estimated number-right score which represents status with respect to achievement on a particular criterion set of test items). The standardized T score is a transformation of the IRT theta (ability) estimate, rescaled to a mean of 50 and standard deviation of 10.
##### 2. [F1TXMSTD] F1 math test standardized score 
####### Description: Math standardized T Score. The standardized T score provides a norm-referenced measurement of achievement, that is, an estimate of achievement relative to the population (spring 2004 12th-graders) as a whole. It provides information on status compared with peers (as distinguished from the IRT-estimated number-right score which represents status with respect to achievement on a particular criterion set of test items). Although the T score is reported for all F1 in-school responding students (including transfer students), regardless of grade level, the comparison group for standardizing is the 12th grade population. The standardized T score is a transformation of the IRT theta (ability) estimate, and has a mean of 50 and standard deviation of 10 for the weighted subset of 12th-graders in the sample.
```


```{r data cleaning, message=FALSE}
# Data Cleaning 
#retrieve data
#els <- read_csv("./data/els_02_12_byf3pststu_v1_0.csv")
#select columns 
#els <- els %>% dplyr::select(STU_ID, BYSEX, BYRACE, BYMOTHED, BYTXMSTD, F1TXMSTD)
#save the revised (cleaned) data to csv
#write.csv(els,"./data/els_cleaned.csv", row.names = FALSE)

els <- read_csv("./data/els_cleaned.csv")

#replace missing data code to NA
els$BYSEX <- na_if(els$BYSEX, -4)
els$BYSEX <- na_if(els$BYSEX, -8)
els$BYRACE <- na_if(els$BYRACE, -4)
els$BYRACE <- na_if(els$BYRACE, -8)
els$BYMOTHED <- na_if(els$BYMOTHED, -4)
els$BYMOTHED <- na_if(els$BYMOTHED, -8)
els$BYMOTHED <- na_if(els$BYMOTHED, -9)
els$BYTXMSTD <- na_if(els$BYTXMSTD, -8)
els$F1TXMSTD <- na_if(els$F1TXMSTD, -8)

#remove if the row doesn't have both BY and F1 math scores
els <- els %>% 
  filter(!is.na(BYTXMSTD) | !is.na(F1TXMSTD))

#rename
els <- els %>%
  mutate(BYSEX = dplyr::recode(BYSEX, 
                          `1` = "Male", 
                          `2` = "Female"),
         BYRACE = dplyr::recode(BYRACE, 
                          `1` = "Native American/Alaskan", 
                          `2` = "Asian", 
                          `3` = "Black", 
                          `4` = "Hispanic (no race specified)", 
                          `5` = "Hispanic (specified)", 
                          `6` = "More than one race, non-Hispanic", 
                          `7` = "White, non-Hispanic"),
         BYMOTHED = dplyr::recode(BYMOTHED,
                          `1` = "Did not finish high school",
                          `2` = "Graduated from high school or GED",
                          `3` = "Attended 2-year school, no degree",
                          `4` = "Graduated from 2-year school",
                          `5` = "Attended college, no 4-year degree",
                          `6` = "Graduated from college",
                          `7` = "Completed Master's degree or equivalent",
                          `8` = "Completed PhD, MD, other advanced degree"))


#rename columns to use pivot_longer
colnames(els)[colnames(els) %in% c("BYTXMSTD", "F1TXMSTD")] <- c("Base", "Follow-up")

els_longer <- els %>% 
  pivot_longer(
    cols = c('Base', 'Follow-up'),
    names_to = "YEAR",
    values_to = "MATH"
  )

els_wider_by <- els %>% 
  pivot_wider(
    id_cols = !'Follow-up',
    names_from = BYRACE,
    values_from = c(Base)
  )

els_wider_f1 <- els %>% 
  pivot_wider(
    id_cols = !Base,
    names_from = BYRACE,
    values_from = c('Follow-up')
  )

```
## Visualization 1

```{r visualization 1, message=FALSE, warning=FALSE}
vis1Data <- els_longer %>% 
  mutate(YEAR = factor(YEAR,
                       levels = c("Follow-up",
                                 "Base"))) %>%
  filter(!is.na(BYSEX)) %>%
  ggplot(aes(x=MATH,y=YEAR,fill=YEAR)) +
    geom_col(position="dodge", show.legend = FALSE) +
    facet_wrap(~ BYSEX,ncol=1) +
    labs(x="Math Scores",
         y="Year",
         title="Student Math Scores",
         subtitle="by year and sex"
         ) +
    scale_fill_manual(values = c("maroon", "gold")) +
  theme_light()
vis1Data
```

## Visualization 2

```{r visualization 2, message=FALSE, warning=FALSE}
vis2Data <- els_longer %>% 
  mutate(YEAR = factor(YEAR,
                       levels = c("Follow-up",
                                  "Base"))) %>%
  filter(!is.na(BYRACE)) %>%
  ggplot(aes(x=MATH,y=YEAR,fill=YEAR)) +
    geom_col(position="dodge", show.legend = FALSE) +
    facet_wrap(~ BYRACE,ncol=1) +
    labs(x="Math Scores",
         y="Year",
         title="Student Math Scores",
         subtitle="by year and race"
         ) +
    scale_fill_manual(values = c("maroon", "gold")) +
  theme_light()
vis2Data

# Alternate graph combining Visualization 1 & 2? Maybe easier that way?

vis2DataAlternate <- els_longer %>% 
  mutate(YEAR = factor(YEAR,
                       levels = c("Follow-up",
                                  "Base"))) %>%
  mutate(BYSEX = factor(BYSEX,
                       levels = c("Male",
                                  "Female"))) %>%
  filter(!is.na(BYRACE)) %>%
  ggplot(aes(x=MATH,y=YEAR,fill=BYSEX)) +
    geom_col(position="dodge") +
    facet_wrap(~ BYRACE,ncol=1) +
    scale_fill_discrete(breaks=c('Male', 'Female')) +
    labs(x="Math Scores",
         y="Year",
         fill = "Sex",
         title="Student Math Scores",
         subtitle="by year and race, separated by sex"
         ) +
    scale_fill_manual(values = c("maroon", "gold")) +
  theme_light()
vis2DataAlternate
```

## Visualization 3

### Distribution plots
These are simple distribution plots by Race, Sex, and Mother's education for year 1 and year 2 for standardized math scores.

```{r visualization 3, message=FALSE, warning=FALSE}
# Fixed the names to look better.
els_viz <- els_longer %>%
  mutate(RACE = dplyr::recode(BYRACE, "Native American/Alaskan" = "Native American\n /Alaskan",
                         "Asian" = "Asian", 
                         "Black" = "Black", 
                         "Hispanic (no race specified)" = "Hispanic", 
                         "Hispanic (specified)" = "Hispanic\n (Race specified)", 
                         "More than one race, non-Hispanic" = "2+ races\n non-Hispanic",
                         "White, non-Hispanic" = "White\n non-Hispanic"),
         MOTHED = dplyr::recode(BYMOTHED, 
                           "Did not finish high school" = "Did not finish\n high school",
                           "Graduated from high school or GED" = "Graduated high\n school or GED",
                           "Attended 2-year school, no degree" = "Attended 2-year school\n no degree",
                           "Graduated from 2-year school" = "Graduated 2-year\n school",
                           "Attended college, no 4-year degree" = "Attended college\n no degree",
                           "Graduated from college" = "Graduated college",
                           "Completed Master's degree or equivalent" = "Master's degree",
                           "Completed PhD, MD, other advanced degree" = "PhD, MD,other\nadvanced degree")) %>%
  mutate(RACE = factor(RACE, levels = c("White\n non-Hispanic",
                                        "Black",
                                        "Hispanic",
                                        "Hispanic\n (Race specified)",
                                        "Asian",
                                        "Native American\n /Alaskan",
                                        "2+ races\n non-Hispanic")),
         MOTHED = factor(MOTHED, levels = c("Did not finish\n high school",
                                            "Graduated high\n school or GED",
                                            "Attended 2-year school\n no degree",
                                            "Graduated 2-year\n school",
                                            "Attended college\n no degree",
                                            "Graduated college",
                                            "Master's degree",
                                            "PhD, MD,other\nadvanced degree")))


# A plot of the distribution of math scores by race in year 1 and follow up  
els_viz %>%
  filter(!is.na(MATH) & !is.na(RACE)) %>%
  ggplot(aes(x = MATH)) +
  geom_histogram(col='black',fill='white')+
  theme_minimal() +
  xlab("Math Scores") +
  xlim(10,90)+
  facet_wrap( ~ RACE + YEAR, nrow = 2, ncol=7)+
  theme(strip.background =element_rect(fill="white"))

# A plot of the distribution of math scores by sex in year 1 and follow up.
els_viz %>%
  filter(!is.na(MATH) & !is.na(RACE) & !is.na(BYSEX)) %>%
  ggplot(aes(x = MATH)) +
  geom_histogram(col='black',fill='white')+
  theme_minimal() +
  xlab("Math Scores") +
  xlim(10,90)+
  facet_wrap( ~ BYSEX + YEAR, nrow = 1, ncol=4)+
  theme(strip.background =element_rect(fill="white"))

# A Plot of distribution of math scores by mother's education for year 1 and follow up. 
els_viz %>%
  filter(!is.na(MATH) & !is.na(MOTHED)) %>%
  ggplot(aes(x = MATH)) +
  geom_histogram(col='black',fill='white')+
  theme_minimal() +
  xlab("Math Scores") +
  xlim(10,90)+
  facet_wrap( ~ MOTHED + YEAR, nrow = 2, ncol=8)+
  theme(strip.background =element_rect(fill="white"))

# Honestly I don't love the visualization, so let's look at a box plot of the data.

# How about a density plot?
```

### Boxplots
A boxplot of the standardized math scores by Race, Sex, and Mother's education.  

```{r visualization 4, message=FALSE, warning=FALSE}
# Boxplot of math scores by Race and Sex separated by Year
els_viz %>%
  filter(!is.na(MATH) & !is.na(RACE)) %>%
  ggplot(aes(x= RACE, y=MATH)) +
  geom_boxplot(aes(fill = RACE), show.legend = FALSE)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_viridis_d(option = 'plasma')+
  theme_minimal()+
  facet_wrap(~BYSEX)+
  labs(x = "", 
       y = "Math Scores",
       title = "Math Score by Race and Year",
       subtitle = "Separated by sex")+
  coord_flip()

# Boxplot of math scores by Mother's education separated by Year
els_viz %>%
  filter(!is.na(MATH) & !is.na(MOTHED)) %>%
  ggplot(aes(x= MOTHED, y=MATH)) +
  geom_boxplot(aes(fill = MOTHED), show.legend = FALSE)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  coord_flip()+
  scale_fill_viridis_d(option = 'mako')+
  theme_minimal()+
  facet_wrap(~YEAR)+
  labs(x = "", 
       y = "Math Scores",
       title = "Math Score by Mother Education and Year")

# Boxplot of math scores by sex separated by year
els_viz %>%
  filter(!is.na(MATH) & !is.na(BYSEX)) %>%
  ggplot(aes(x= BYSEX, y=MATH)) +
  geom_boxplot(aes(fill = BYSEX), show.legend = FALSE)+
  scale_fill_viridis_d()+
  theme_minimal()+
  facet_wrap(~YEAR)+
  labs(x = "", 
       y = "Math Scores",
       title = "Math Score by Sex and Year")
```

### Density Plots

Finally let's examine the data using density plots.
```{r visualization 5, message=FALSE, warning=FALSE}
# Density plot of math scores by Race and Sex.
els_viz %>%
  filter(!is.na(MATH) & !is.na(RACE)) %>%
  ggplot(aes(x = MATH, y = RACE))+
  geom_density_ridges(aes(fill = RACE), alpha=0.5)+
  scale_fill_viridis_d(option = 'plasma')+
  theme_minimal()+
  theme(legend.position = "none")+
  facet_wrap(~BYSEX)

# Denisty plot of math scores by Mother education
els_viz %>%
  filter(!is.na(MATH) & !is.na(MOTHED)) %>%
  ggplot(aes(x = MATH, y = MOTHED))+
  geom_density_ridges(aes(fill = MOTHED), alpha=0.5)+
  scale_fill_viridis_d()+
  theme_minimal()+
  theme(legend.position = "none")+
  labs(x = "Math Score",
       y = "Mother's Education Level")

# Density plot of math scores by mother education separated by year
els_viz %>%
  filter(!is.na(MATH) & !is.na(MOTHED)) %>%
  ggplot(aes(x = MATH, y = MOTHED))+
  geom_density_ridges(aes(fill = MOTHED), alpha=0.5)+
  scale_fill_viridis_d()+
  theme_minimal()+
  theme(legend.position = "none")+
  labs(x = "Math Score",
       y = "Mother's Education Level")+
  facet_wrap(~BYSEX, nrow = 1)

```

## Descriptive Statistics

```{r descriptive, message=FALSE}
#Created table by Race 
By_Race <- els_longer %>%
  group_by(BYRACE) %>%
  summarize(race_n = n(), 
            mean_math = mean(MATH, na.rm = TRUE), 
           sd_math = sd(MATH, na.rm = TRUE))
#Use describeBy to get kurtosis and skew 
describeBy(els_longer$MATH, els_longer$BYRACE)

By_Race %>%
  kbl(caption = "Math Descriptives by Race", 
      digits = 2, 
      col.names = c("Reported Race", "n", "Mean ", "SD")) %>%
  kable_classic()

#table for Mothers Education 
By_MotherED <- els_longer %>%
  group_by(BYMOTHED) %>%
  summarize(mothed_n = n(), 
            mean_math = mean(MATH, na.rm = TRUE), 
            sd_math = sd(MATH, na.rm = TRUE))

describeBy(els_longer$MATH, els_longer$BYMOTHED)

By_MotherED %>%
  kbl(caption = "Math Descriptives by Mother's Education Level", 
      digits = 2, 
      col.names = c("Mother's Education Level", "n", "Mean ", "SD")) %>%
  kable_classic()
 #table for year            
By_Year <- els_longer %>%
  group_by(YEAR) %>%
  summarize(year_n = n(), 
            mean_math = mean(MATH, na.rm = TRUE), 
            sd_math = sd(MATH, na.rm = TRUE))

describeBy(els_longer$MATH, els_longer$YEAR)

By_Year %>%
  kbl(caption = "Math Descriptives by Year", 
      digits = 2, 
      col.names = c("Year", "n", "Mean ", "SD")) %>%
  kable_classic()
#Table by sex 
By_Sex <- els_longer %>%
  group_by(BYSEX) %>%
  summarize(sex_n = n(), 
            mean_math = mean(MATH, na.rm = TRUE), 
            sd_math = sd(MATH, na.rm = TRUE))

 describeBy(els_longer$MATH, els_longer$BYSEX)

By_Sex %>%
  kbl(caption = "Math Descriptives by Sex", 
      digits = 2, 
      col.names = c("Reported Sex", "n", "Mean ", "SD")) %>%
  kable_classic()
```

## More Statistics
### ANOVA Results
#### ANOVA Math scores by Mother Education and Race

```{r ANOVA, warning=FALSE, message=FALSE, fig.width= 14, fig.height= 80}

library(car)
math_mod <- lm(MATH ~ 1 + MOTHED*RACE, data = els_viz)
Anova(math_mod, type = 3)

# Too many comparisons. Instead we are going to look at one way anova comparisons
aovRaceMothed <- aov(MATH ~ RACE*MOTHED, data = els_viz)
tukey_all <- TukeyHSD(aovRaceMothed, conf.level = 0.95)

```

### Post Hoc Tukeys HSD Race and Mother's education
#### Tukeys HSD Race Pairs
```{r pairwise comparison viz, fig.height= 50, fig.width= 12}
aovRace <-  aov(MATH ~ RACE,data=els_viz)
tukey_race <- TukeyHSD(aovRace,conf.level = 0.95)

# Visualization of Tukeys HSD pairwise comparisons by race
plot(tukey_race, col = "brown")

# Let's look if Mother's education is a predictor. Did not finish high school is the reference group
#contrasts(els_viz$MOTHED)

```

#### Tukeys HSD Mother's Ed pairs
```{r, fig.height= 100, fig.width= 12}
aovMothed <-  aov(MATH ~ MOTHED,data=els_viz)
tukey_mothed <- TukeyHSD(aovMothed,conf.level = 0.95)

# Visualization of Tukeys HSD pairwise comparisons by race
plot(tukey_mothed, col = "red")

# Let's look if Mother's education is a predictor. Did not finish high school is the reference group
#contrasts(els_viz$MOTHED)
```

### Student's T test Analysis
#### Difference in scores between year 1-2

```{r t_test}
#str(els_viz)
# Paired student's t test to examine if means from year 1 and year 2 are significantly different

# Created a small data set of just years and scores plus student id
els_byyear <- els_viz %>%
  group_by(YEAR)%>%
  filter(!is.na(MATH) | !is.na(YEAR)) %>%
  select(STU_ID, YEAR, MATH)%>%
  pivot_wider(names_from = YEAR,
              values_from = MATH) %>%
  rename("Follow" = "Follow-up")

# Paired student's t test to examine if the mean from the follow up is significantly greater than the base year mean

# Paired student's t test to examine if means from base year and follow up are significantly different.
t.test(els_byyear$Base, els_byyear$Follow, paired = TRUE)

# Visualization of the means by year. The t test suggests that there is a significant difference in scores by year, despite the means being similar, unfortunately, the mean of the follow up is significantly less than the base year. I was thinking we could focus on visualizing the data by race and mother's education and sex rather than comparing year 1 to year 2 data.
ggwithinstats(data = els_viz, x = YEAR, y = MATH, 
              type = "parametric", 
              centrally.plotting = TRUE, 
              pairwise.display = "s", 
              point.path = FALSE,
              point.args = aes(size = 0, alpha = 0.2),
              results.subtitle = FALSE,
              alternative = "greater")

```

### Regression Analysis
```{r regression, warning=FALSE, fig.height=50, fig.width=12}
# First let us look if Race is a predictor. I will set white as the reference group since it is the largest group. 
#contrasts(els_viz$RACE)
# Looks like I set white as the reference earlier.  

mod_race <- lm(MATH ~ 1 + RACE, els_viz)
#tab_model(mod_race)
# This just tells us that all groups are significantly different from each other
# I think we could report pairwise comparisons

race_pairs <- pairwise.t.test(els_viz$MATH, els_viz$RACE, p.adjust.method = "bonf")
race_pval <- race_pairs$p.value %>%
  round(digits = 3)

options(knitr.kable.NA = "")
race_pval %>%
  kbl(caption = "p -values of Math Score by Race",
      digits = 3) %>%
  kable_classic()
  
# Based on this, most groups are significantly different **except** for black/Native, Hispanic/Hispanic (Race specified), Native/Hispanic, Native/Hispanic (Race specified)

mod_mothed <- lm(MATH ~ 1 + MOTHED, els_viz)
#tab_model(mod_mothed)
# This just tells us that groups are significantly different from each other
# I think we could report pairwise comparisons

mothed_pairs <- pairwise.t.test(els_viz$MATH, els_viz$MOTHED, p.adjust.method = "bonf")
mothed_pval <- mothed_pairs$p.value %>%
  round(digits = 3)

options(knitr.kable.NA = "")
mothed_pval %>%
  kbl(caption = "p -values of Math Score by Mother's Education level",
      digits = 3) %>%
  kable_classic()


```


#### Mother Education / Race Regression  
I am not sure how to do a regression analysis of all three variables, but I can look at the regression analysis of math scores based on race/ethnicity and mother's education.   

```{r regression mother edu and race, fig.width=20}
mod_moth_race <- lm(MATH ~ 1 + MOTHED*RACE, data = els_viz)
#summary(mod_moth_race)

#tab_model(mod_moth_race)
```

#### Mother Edu / Race Regression Visualization  
Now let's try to visualize this, it might get messy.  

```{r, fig.width= 16, fig.height=14}
# Visualization of regression models for mother's education and race
#install.packages("interactions")
cat_plot(mod_moth_race, pred = MOTHED, modx = RACE, geom = "line", interval = FALSE,vary.lty = TRUE)

```

#### Mother ED / Race Bar graph
This allows for the error bars to be included but crowds the data and makes analyzing the data across groups more difficult.

```{r, fig.width= 20, fig.height=10}

cat_plot(mod_moth_race, pred = MOTHED, modx = RACE, geom = "bar")

```


